---
title: "lab_07"
author: "Derek Willis/Rob Wells"
date: "2025-10-21"
output: html_document
---


Lab07 - API Keys and Tidycensus


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Security and API keys

You'll need your Census API Key. Here's how to do that securely:

r_environ is a hidden file that can securely store API keys
```{r}
usethis::edit_r_environ()
# a new text file opens in RStudio
```

1. Add your API keys in the r_environ file
CENSUS_API_KEY=XXXXX
OPENAI_API_KEY=XXXX
etc....

2. Save the r_environ() file and RESTART R

# Hide your API Key from GitHub

```{r}
# Check if .gitignore exists in your project
file.exists(".gitignore")
```

Assuming you don't have a .gitignore file, then create one
1. RStudio:
- File → New File → Text File

2. ## Add These Lines to Your `.gitignore`
# R files
.Rhistory
.RData
.Rproj.user/
.Renviron
.Rprofile

# Environment files
.env
*.env

# API keys and secrets
*secret*
*password*
*api_key*

3. save the file a .gitignore in the root directory of your project
yes, you have to precede the file with a dot: .gitignore

4. Veryify it is working

```{r}
system("git status")
```


## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
census_api_key("40c2313fc1a231b500335ab147c5d6764b80fdd4")
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2023 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2023 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
ACS <- load_variables(2023, "acs5", cache = TRUE)
View(ACS)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2023 5-year estimates? Hint: look at the variables for ACS in 2023 and find the variable for median age. **A1**

```{r}
md <- get_acs(geography = 'county',
              variables = c(medage = 'B06002_001'),
              state = 'MD',
              year = 2023)
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe created in Q1 to answer this, but you must write code to display only those counties with a median age of at least 50. **A2**

```{r}
md |>
  filter(estimate >= 50)
```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  Load the Montgomery County 911 overdose calls data.
```{r}
overdose_calls <- read_csv("/Users/rebeccaheimbrock/Desktop/jour772/lab_07/data/montgomery_2022_overdoses.csv")
```

2.  Create a new dataframe with the total number of calls for each zip code.
```{r}
total_call_no_by_zip <- overdose_calls |> 
  group_by(zip) |> 
  summarise(
    total_calls=n()
  )
```

3.  Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one). Then get from the ACS data from all Maryland zip codes using that variable - you will need to specify an earlier year to get the data. B09021_001
```{r}
eighteen_and_up <- get_acs(geography = 'county',
              variables = c(total_18_and_up = 'B09021_001'),
              state = 'MD',
             )
```
```{r}
md_zipcodes <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2019)
```


4.  Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
```{r}
#I kept running into an error below that said 'population 2019 not found' - I asked Claude how to edit my existing code to grab population per year and make it a new table. The below code is modified per Claude's response.
# Get data for each year
pop_2015 <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2015) |>
  select(GEOID, Population2015 = estimate)

pop_2016 <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2016) |>
  select(GEOID, Population2016 = estimate)

pop_2017 <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2017) |>
  select(GEOID, Population2017 = estimate)

pop_2018 <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2018) |>
  select(GEOID, Population2018 = estimate)

pop_2019 <- get_acs(geography="zcta", variables = "B09021_001", state='MD', year=2019) |>
  select(GEOID, Population2019 = estimate)

# Join them all together
ods_age_and_zip <- md_zipcodes |>
  left_join(pop_2015, by='GEOID') |>
  left_join(pop_2016, by='GEOID') |>
  left_join(pop_2017, by='GEOID') |>
  left_join(pop_2018, by='GEOID') |>
  left_join(pop_2019, by='GEOID')
```

```{r}

```

5.  Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.
```{r}
ods_age_and_zip |> 
  mutate(change = ((Population2019 - Population2015) / Population2015) * 100) |> 
  arrange(desc(change))
```


Which zip code has the highest rate of calls? Which has the lowest?

**A3** The zip with the highest rate of calls is Massey, Maryland. Massey has 

```{r}

```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4**

**Write your answer here:**

