---
title: "lab_03"
author: "Bex Heimbrock"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008) |> 
clean_names()
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |> 
clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. 

Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: 51 obs of 8 variables!

```{r}
pg_schools <- payments |> 
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS|PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS|PG COUNTY PUBLIC SCHOOL|PRINCE GEORGES PUBLIC SCHOOLS"))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The 2020 fiscal year had the largest amount of spending. This is probably because of added costs of online learning, like contracts with Blackboard or Zoom, and maybe hazard pay? I think it's interesting that spending has decreased noticeably in 2025. Is this because of the MD deficit?

```{r}
pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(total_amt=sum(amount, nam.rm=TRUE)) |> 
  arrange(desc(total_amt)) 
```

Q3. This question is in two parts. 

Part 1: Code: You're interested in examining the vendors paid for capital projects at the University of Maryland. Write code that does the following: 
--Filters payments data to capital projects at the University of Maryland
--Produce a list of vendors and the total amount they received. 

```{r}
cap_projects_md <- payments |> 
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND"))
```

```{r}
cap_projects_md |> 
   group_by(vendor_name, vendor_zip) |> 
  summarise(total_amt=sum(amount, nam.rm=TRUE)) |> 
  arrange(desc(total_amt)) 
```
Part 2: Provide a written answer, not code.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? 
--Provide at least three examples of vendor names that are similar and should be combined into one. 
--Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?


A3. We would need to create a code that combines the different spellings of names into one name - this would be a stringer detect function. We will also need a code that puts the zip codes of companies with the same or similar names into the same row -- we would do this the same way we did in the pre-lab. These are the instructions from the pre-lab: "We're going to write code that tells R to make a new column for our zips, keeping the first five digits on the left, and get rid of anything after that by using `mutate()` in concert with `str_sub()`, from the `stringr` package."

Three examples of vendor names are ... Whiting Turner Contracting Co & Whiting Turner Contracting Company, Plano Coudon LLC & Plano Coudon, LLC, and DISYS SOLUTIONS INC with 2 different zip codes. 

```{r}
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. 

Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
Then create a total amount of money for each fiscal year. 

Write a sentence summarizing how stem cell research funding has changed over time. Be specific with the totals.

A4: In 2009, stem cell research was funded by the state of MD to the amount of 14,643,722. In 2021, the amount was 4,832,226. This shows that the amount has decreased drastically. We can ask more questions here -- is the decrease for political reasons, such as the moral panic about CRISPR etc? Is more funding coming from outside sources, instead?

```{r}
stem_funding <- md_grants_loans |> 
  filter(str_detect(description, "Stem Cell|STEM CELL|stem cell")) |> 
   group_by(fiscal_year) |> 
   summarise(total_amt=sum(amount, nam.rm=TRUE)) |> 
   arrange(desc(total_amt)) 
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:There are 33 grantees from that zip code. If I wanted one record per grantee, I would need to use mutate and case when to add a column combining the grantees with same or similar names. I am very interested by the solar panel grant given to IKEA college park. Why couldn't the massive company have paid for it themselves?

```{r}
md_grants_loans |> 
  filter(str_detect(zip_code, "20740")) |> 
  group_by(grantee) |> 
  summarise(total_amt=sum(amount, nam.rm=TRUE)) |> 
  arrange(desc(total_amt)) 
```
